---
title: "W203 Lab 3"
author: "Mohammad Jawad Habib"
date: "April 3, 2016"
output: pdf_document
---

# Part 1: Multiple Choice
Q1 : b
Q2 : b, c 
Q3 : b
Q4 : c
Q5 : d
Q6 : a
Q7 : b
Q8 : b

# Part 2: Test Selection
Q9 : e, Chi-square test
Q10: d, ANOVA
Q11: a, t-test
Q12: b, Pearson correlation
Q13: d, Chi-square test


# Part 3: Data Analysis and Short Answer
```{r}
# setwd("W203 Week 12/Lab 3")
load("GSS.Rdata")
```

## 14. Task 1: Chi-Square Test on Marital Status and Political Orientation

###A. Null and Alternative Hypothesis
H0: marital status and political orientation are independent
H1: marital status and poliical orientation are not independent (that is, knowning about marital status can help predict the political orientation).

We can assume that the various marital statuses are independent of each other i.e. one person is not married and widowed at the same time. We can assume independence for political categorization as well.

We will also see below that none of the expected frequencies are below 5.

```{r}
mp <- data.frame(GSS$marital, GSS$politics)

# remove the value coded as "NA" from "marital"
mp$GSS.marital[mp$GSS.marital == "NA"] <- NA

# remove "NA" as a factor from marital
mp$GSS.marital <- factor(mp$GSS.marital, 
                         c("married", "widowed", "divorced", "separated", 
                           "never married"))

# remove all rows that have NA for "marital" or "politics"
mp <- mp[complete.cases(mp),]

# run the Chi Square test
mpcs <- chisq.test(table(mp))

# see that no expected frequencies are below 5
mpcs$expected

```

###B. Test Statistics and p-value
We get a test statistic and p-value as follows:
```{r}
# test statistic
mpcs$statistic

# p-value
mpcs$p.value
```

Given the p-value above we can reject the null hypothesis. We can say that marital status does seem to be related to political orientation.

###C. Effect Size Calculation
We will use Cramer's V for the effect size.
```{r}
cv <- sqrt(mpcs$statistic / (length(mp$GSS.marital)*min(nrow(mpcs$observed) - 1, ncol(mpcs$observed) - 1)))
names(cv) <- "Cramer's V"
cv
```

###D. Interpretation
Our Chi-Square test reveals that "politics" is significantly related to "marital" status: $$\chi^2 (16) = 44.225, p < 0.01$$. The contingency table is shown below.
```{r}
mpcs$observed
```

We can also see from the standardized residuals which ones are significant outside of +/- 1.96 (p < 0.05).
```{r}
mpcs$stdres > 1.96 | mpcs$stdres < -1.96
```
Being "married" is significantly related with "Conservative" political view.
"widowed" is not significantly related with any political view.
"divorced" is significantly related with "Conservative".
"separated" is significantly related with "Moderate" and "Conservative".
"never married" is significantly related with "Liberal" and "Conservative".


## 15. Task 2: Pearson Correlation on Age when Married and Hours of TV watched

###A. Null and Alternative Hypothesis
H0: there is no relationship between agewed and tvhours (r = 0)
H1: there is a positive or negative relationship between agewed and tvhours

###B. Test Statistics and p-value
GSS website is not very clear on how missing values in `agewed` and `tvhours` are coded. For `agewed` let's assume that 0 and 99 are dummy values. for `tvhours` we can assume that anything greater than 24 is a dummy value because there are only 24 hours in the day.

Note: our sample size is large so we can assume normality of our sampling distribution. This assumption is required for establishing whether correlation coefficient is significant.

```{r, warning=FALSE, message=FALSE}
at <- data.frame(GSS$agewed, GSS$tvhours)
at <- at[!(at$GSS.agewed %in% c(0, 99)) & at$GSS.tvhours <= 24,]

library(Hmisc)

atpc <- rcorr(as.matrix(at))
atpc

cor.test(at$GSS.agewed, at$GSS.tvhours)
```


From the above, we get a test statistic (r) of -0.03 and a p-value of 0.3009. And we get a 95% confidence interval that passes through zero (-0.087, 0.027)

###C. Interpretation
The results indicate that agewed is not correlated with tvhours. We also get $$r^2 = 9e-04$$ which means that only 0.09% of variability is shared between agewed and tvhours.

I ran `shapiro.wilk` on `agewed` and `tvhours` and the two are not normally distributed. We can also check this by `hist`. Therefore, I'm going to run a Kendall's Tau. Note: we relied on a large sample size before to assume normality but we don't need that assumption with the non-parametric Kendall's Tau. I did not use Spearman because of ties in ranked data and I also did not use bootstrapping (which would be trivial anyways).
```{r}
cor.test(at$GSS.agewed, at$GSS.tvhours, method = "kendall")
```
Even though p < 0.01, the tau is -0.06 which shows that there may be a weak negative relationship between agewed and tvhours.

## 16. Task 3: Wilcox Rank-Sum on Marital Status and Number of Children for 23 year olds

We will remove the observation where marital value is set to "NA" and use the rest of the data for our analysis. We will also subset the data for age==23 aftewards (we could do this before too). 
```{r}
mc <- GSS[, c("age", "marital", "childs")]
mc$marital[mc$marital == "NA"] <- NA
mc$marital <- factor(mc$marital, 
                     c("married", "widowed", "divorced", "separated", 
                       "never married"))
mc <- mc[complete.cases(mc),]

mc$married <- ifelse(mc$marital == "married", 1, 0)
# mc$married <- factor(mc$married, c("married", "not married"))

# just keep those who are 23
mc <- mc[mc$age == 23,]


```

### A. Mean of Married Variable
```{r}
sum(mc$married) / length(mc$married)

```
The proportion of observations coded married = 1 is 0.286 in our subset.

### B. Null and Alternative Hypothesis
H0: median number of children is the same for married and non-married 23-year olds
H1: median number of children is not equal for married and non-married 23-year olds

### C. Test Statistic and p-value
```{r}
# mc$married <- factor(mc$married, c("married", "not married"))
mcwrs <- wilcox.test(childs ~ as.factor(married), data = mc, exact = FALSE)
mcwrs
```
From the above test, we see that W = 19 and p = 0.000266 (p < 0.01).

### D. Calculate Effect Size
```{r}
z <- qnorm(mcwrs$p.value/2)
r <- z/sqrt(length(mc$married))
r
```
We see from the above that the effect size r is -0.689 which is conventionally considered a large effect (above 0.5).

###E. Interpretation
Our results show that number of children in married 23-year olds (Mdn = 1) differed significantly from unmarried 23-year olds (Mdn = 0), W=19, p = 0.000266, r = -0.689. That is married 23-year olds had significantly more children than unmarried ones.

## 17. Task 4: ANOVA on Religious Affiliation and Age When Married

### A. Null and Alternative Hypothesis
H0: mean age when married is the same across all religious affiliation
H1: mean age when married is NOT the same across all religious affiliation

### B. Test Statistic and p-value
We will remove the observations where agewed equals 0 or 99. We will also remove the observations where relig is NA or DK (per the GSS website).
```{r, warning=FALSE, message=FALSE}
ar <- data.frame(GSS$agewed, GSS$relig)
ar <- ar[!(ar$GSS.agewed %in% c(0, 99)) & !(ar$GSS.relig %in% c("NA", "DK")),]

ar$GSS.relig <- factor(ar$GSS.relig, 
                       c("Protestant", "Catholic", "Jewish", "None", "Other"))
names(ar) <- c("agewed", "relig")

require(ggplot2)

arp <- ggplot(ar, aes(relig, agewed)) + 
  stat_summary(fun.y = mean, geom = "line", size = 0.5, aes(group = 1), 
               colour = "#990000") + 
  stat_summary(fun.data = mean_cl_boot, geom = "errorbar", 
               width = 0.2, size = 0.5, colour = "#990000") + 
  stat_summary(fun.y = mean, geom = "point", size = 3, colour = "#FF6633")
arp

```
The errorbars for protestants do not overlap with errorbars of catholic, jewish and other.

```{r, warning=FALSE, message=FALSE}
require(pastecs)
by(ar$agewed, ar$relig, stat.desc)

require(car)
leveneTest(ar$agewed ~ ar$relig, center = median)

by(ar$agewed, ar$relig, shapiro.test)

```
The non-significant Levene Test above shows that we have homogeneity of variance.
We do not have normally distributed groups for Protestant, Catholic and None. We do have normally distributed groups for Jewish and Other.

Section 10.3.2 of our text book says that F-statistic controls Type I error well under conditions of non-normality. Power of F-statistic is also relatively unaffected by non-normality per our text book when group sizes are equal. Unfortunately, that is not the case so we will stick with a robust ANOVA as described in section 10.6.6.3 of our text book.

```{r, warning=FALSE, message=FALSE}
require(WRS)

arwide <- unstack(ar, agewed ~ relig)

# resample to make groups sizes equal
arwide <- data.frame(lapply(arwide, sample, size  = 30, replace = TRUE))

# t1way
levels(arwide)

WRS::med1way(arwide)
WRS::t1waybt(arwide, nboot = 2000)

```
We have a non-significant result for median age by religion and a marginally significant result for mean.

Using the results from t1waybt above, we get a test statistic (F) of 3.380 and p.value of 0.042. From the med1way we get a test statistic (F) of 1.69 and p.value of 0.09.

### C. Statistical Differences Between Individual Pairs
Since we did not have any directional hypothesis beforehand, we will run post-hoc tests.

```{r}
arpt <- pairwise.t.test(ar$agewed, ar$relig, paired = FALSE, p.adjust.method = "bonferroni")
arpt
```

Result table (`pairwise.t.test`) above shows that Catholic-Protestant (p < 0.01), Jewish-Protestant (p < 0.05) and Other-Protestant (p < 0.05) had significantly different age at wedding.

### D. Evaluate Hypothesis
We need to know the p-value and confidence interval to evaluate our hypothesis.
```{r}
amodel <- aov(agewed ~ relig, ar)
TukeyHSD(amodel)
```

We can see from the above that Protestant-Catholic (p < 0.01), Protestant-Jewish (p < 0.05), Protestant-Other (p < 0.05) have 95% confidence intervals that do not pass through zero.

We can also calculate the desired effect sizes (Cohen's D).
```{r}
# Effect Size Calculation
# Protestant-Catholic
d <- ar[ar$relig %in% c("Protestant", "Catholic"),]
d$relig <- factor(d$relig, c("Protestant", "Catholic"))
cohensD(agewed ~ relig, data = d)

# Protestant-Jewish
d <- ar[ar$relig %in% c("Protestant", "Jewish"),]
d$relig <- factor(d$relig, c("Protestant", "Jewish"))
cohensD(agewed ~ relig, data = d)

# Protestant-Other
d <- ar[ar$relig %in% c("Protestant", "Other"),]
d$relig <- factor(d$relig, c("Protestant", "Other"))
cohensD(agewed ~ relig, data = d)
```
As calculated above, Cohen's D indicates a significant effect size for mean age (higher) of Catholic, Jewish and Others vs. that of Protestants.

