---
title: "W203 Final"
author: "Mohammad Jawad Habib"
date: "April 21, 2016"
output: pdf_document
---

# Part 1: Multiple Choice
Q1. b, Ordinary Least Squares regression

Q2. b, All coefficients for each independent variable equal zero

Q3. a, There are statistically significant relationships between each
independent variable and the dependent variable

Q4. b, Maximum number of pushups in 3 minutes (because it might be correlated
with tricep strength which is already in our model)

Q5. d, because we did not ourselves subject people to this 
torment (um, treatment); we assume that the populations of the two CA counties 
do not differ substantially prior to the ISP ban on dating sites

Q6. c, Regression assumptions that have been met

Q7. b, The power of the test (is not a random variable, it's calculated)

Q8. c, Insignificant results are commonly relegated to the file drawer
If this were actually true, only 5% or less of the publish research would 
not be reproducible. So relegating insignificant research to the file drawer 
does not help to explain why more than 5% of the published results appear to 
be type-1 errors

# Part 2. Test Selection
Q9. a, chi-square (use_reddit ~ marital_status), categorilcal, nominal
- t-test cannot be used because it requires numeric data and Gaussian distribution
- Pearson Correlation cannot be used because it requires two variables from a Gaussian distribution
- Wilcoxon signed rank test requires same participants in different measures; limited to two groups

Q10. d, ANOVA (life_quality ~ region) #Kruskal Wallis test would be better
- Pearson correlation cannot be used because it requires two variables from a Gaussian distribution
- Wilcoxon signed rank test requires same participants in different measures; limited to two groups
- Binary logistic regression cannot be used because it requires Bionmial outcomes

Q11. d, ANOVA (flirted_online ~ years_in_relationship)
- Fisher's exact test requires two groups and two outcomes
- t-test cannot be used because it requires numeric data and Gaussian distribution
- Wilcoxon signed rank test requires same participants in different measures; limited to two groups


Q12. a, chi-square (lgbt ~ adults_in_household)
- ANOVA cannot be used because it requires a normally distributed dependent variable
- Pearson correlation cannot be used because it requires two variables from a Gaussian distribution
- Wilcoxon rank sum test requires exactly two groups

Q13. a, Pearson Correlation (totalchildren ~ age)

Q14. b, Wilcoxon Rank-Sum Test

# Part 3. Data Analysis
```{r}
dat <- read.csv("Dating.csv", header = TRUE, stringsAsFactors = FALSE)
```

## 15. OLS Regression

### a. Mean of life_quality
```{r}
unique(dat$life_quality)
# omit "Refused" and "Don't know" and calculate mean

lq <- dat$life_quality[!dat$life_quality %in% c("Refused", "Don't know")]
lq <- lq[!is.na(lq)]
lq <- as.numeric(as.character(lq))
mean(lq) # 2.607079

```

### b. Mean of years_in_relationship
```{r}
unique(dat$years_in_relationship)
# omit "Refused", "Don't know" and " "
# leave "0" in there

yr <- dat$years_in_relationship[!dat$years_in_relationship %in% c("Refused", "Don't know", " ")]
yr <- yr[!is.na(yr)]
yr <- as.numeric(as.character(yr))
mean(yr) # 13.47697
```

### c. Create a subset of data omitting missing values
```{r}
ds1 <- dat[, c("life_quality", "years_in_relationship", "use_internet")]
# apply(ds1, 2, class)
# apply(ds1, 2, levels)


# check what's in use_internet
unique(ds1$use_internet)

# keep only complete cases
ds1 <- apply(ds1, 2, function(x) {ifelse(x %in% c("Refused", "Don't know", " "), NA, x)})
ds1 <- ds1[complete.cases(ds1), ]
nrow(ds1) #1090 cases left
ds1 <- data.frame(ds1, stringsAsFactors = FALSE) # convert to data.frame
ds1[, 1] <- as.numeric(ds1[, 1])
ds1[, 2] <- as.numeric(ds1[, 2])

```

### d. Fit an OLS model: life_quality ~ years_in_relationship
```{r}
lm1 <- lm(life_quality ~ years_in_relationship, data = ds1)

lm1s <- summary(lm1)
lm1s

lm1s$coefficients[1] # intercept is 2.66978, p < 0.001, statistically significant
# Intercept tells us the expected value of life_quality without any independent variable
# So our intercept is very close to the mean value of life_quality which makes sense

lm1s$coefficients[2] # slope is -0.00499, p < 0.05, statistically significant
# however, life_quality only decreases by 0.00498 for 1 year increase in years_in_relationship
# this is a very small effect

lm1s$r.squared # R-squared = 0.00586, i.e. 0.58% of variance in life_quality explained by years_in_relationship

sqrt(lm1s$r.squared) # Pearson's R = 0.07655 (< 0.30) which points to no linear correlation
# This does not appear to be practically significant

cbind("f ratio" = lm1s$fstatistic[1], "critical f value" = qf(c(0.975), 1, 1088)) # F-ratio = 6.41427 
# F-ratio obtained before is greater than the critical value
pf(lm1s$fstatistic[1], lm1s$fstatistic[2], lm1s$fstatistic[3], lower.tail = FALSE) * 2 # p < 0.05


```

### e. Fit a second OLS model
```{r}
lm2 <- lm(life_quality ~ years_in_relationship + use_internet, data = ds1)
lm2s <- summary(lm2)
lm2s

# slope coefficient for use_internet
lm2s$coefficients[3] # slope is -0.40374, p < 0.01, statistically significant
lm2s$adj.r.squared # Adjusted R-squared = 0.02282, i.e. 2.28% of variance in life_quality explained by years_in_relationship & use_internet together
sqrt(lm2s$adj.r.squared) # Pearson's R = 0.15105 (< 0.30) which points to no linear relationship
# This does not appear to be practically significant

cbind("f ratio" = lm2s$fstatistic[1], "critical f value" = qf(c(0.975), 1, 1088)) # F-ratio = 13.71293 
# F-ratio obtained before is greater than the critical value
pf(lm2s$fstatistic[1], lm2s$fstatistic[2], lm2s$fstatistic[3], lower.tail = FALSE) * 2 # p < 0.01

```

### f. F-ratio and p-value between two OLS models
```{r}
lmcompare <- anova(lm1, lm2)
lmcompare

```
Using years_in_relationship and use_internet together improved the fit of the model to the data compared to years_in_relationship alone. F(1, 1087) = 20.894, p < 0.01.

## Q16. Logistic Regression

### a. Odds that a responded has flirted online
```{r}

ds2 <- dat[, c("usr", "flirted_online")]
nrow(ds2)

unique(ds2$flirted_online)
unique(ds2$usr)

ds2 <- apply(ds2, 2, function(x) {ifelse(x %in% c("Refused", "Don't know", " "), NA, x)})
ds2 <- ds2[complete.cases(ds2), ]
nrow(ds2) #1885 cases left
ds2 <- data.frame(ds2, stringsAsFactors = FALSE) # convert to data.frame

library(gmodels)
ct <- CrossTable(ds2$flirted_online)
ct$prop.row[2] / ct$prop.row[1] # odds ratio in sample (flirted / not flirted) = 0.26087

```

### b. Logistic regression: flirted_online ~ usr
```{r}
# convert flirted_online to factor
# ds2$flirted_online <- factor(ds2$flirted_online)
# ds2$flirted_online <- relevel(ds2$flirted_online, ref = "No") # convert no to base

ds2$flirted <- ifelse(ds2$flirted_online == "Yes", 1, 0)
ds2$usr <- factor(ds2$usr) # rural is the base factor

gm <- glm(flirted ~ usr, data = ds2, family = binomial)
gm
gms <- summary(gm)
gms

# Akaike Information Criterion
gms$aic # AIC = 1909
```

### c. Odds of urban user flirting online vs. rural user
```{r}

# use type = "response"" to get probabilities instead of log-odds
pflirt.rural <- predict.glm(gm, newdata = data.frame("usr" = "Rural"), type = "response") 
pflirt.urban <- predict.glm(gm, newdata = data.frame("usr" = "Urban"), type = "response")

pflirt.odds <- pflirt.urban / pflirt.rural
pflirt.odds # Urban user is 1.83 times more likely to flirt than rural user

```