---
title: "W203 Week 8 Lab 2"
author: "Mohammad Jawad Habib"
date: "February 28, 2016"
output: pdf_document
---

# Part 1: Multiple Choice
1. a - Bar Graphs
2. c - $$H_0: \mu = \mu_0; H_a: \mu > \mu_0$$
3. f - none of the above
4. e - Type II error will go up, Power will go down
5. e - Raise the variable to a power greater than 1
6. b - The standard deviation of Berkeley student ages is 2 years
7. d - What is the probability of the data we observe, assuming that the null hypothesis is true
8. c - Assuming your null hypothesis is actually true, and you were to repeat the experiment a large number of times, you would expect a type 1 error 4% of the time. 
9. a - Normal sampling distribution of the test statistic. 
10. f - None of the above


# Part 2: Test Selection

# Part 3: Data Analysis and Short Answer
```{r}
require(knitr)

setwd("~/R/W203/W203 Week 8/")

load("GSS.Rdata")
# kable(head(GSS))
head(GSS)
```

## 13. Data Import and Checking
### 13.a. Examine `agewed`
```{r}
sort(unique(GSS$agewed))
GSS$agewed[GSS$agewed < 18]
GSS$agewed[GSS$agewed > 58]
```
We can assume that `GSS$agewed == 0` and `GSS$agewed == 99` are not reasonable ages to get married. It would also seem that `GSS$agewed < 18` is not a reasonable age to get married. Marriageable age is 18 in most US states and most other countries, and parental consent is required in the US marry younger than that. However, let's assume that there were cases where people got married younger than 18 and consider those ages valid. It must be noted that in some backward countries like Pakistan, children are sometimes "married"" at very young ages, sometimes even at birth, without their consent. That's a horrible practice but we cannot fix that in this assignment. We can only hope that these people will come to their senses soon.

### 13.b. Recode unreasobale `agewed` as NA
```{r}
# As noted before, we will only recode 0 and 99 
# Yes, we're predujiced against the very old when it comes to marriage
GSS$agewed[GSS$agewed == 0] <- NA
GSS$agewed[GSS$agewed == 99] <- NA

# Calculate the mean, ignoring NA
agewed.mean <- mean(GSS$agewed, na.rm = TRUE)
agewed.mean
```
It can be seen that the mean `agewed` is `22.79201` when `0` and `99` are recoded as `NA`.

