---
title: "W203 Week 5 Lab 1"
author: "Mohammad Jawad Habib"
date: "February 7, 2016"
output: pdf_document
---


# Part 2a. Variable Manipulations

## 10. Load Data and calculate gdp_growth

We'll set the working directory to where we saved the "GDP_World_Bank.csv" file and then load it as a `data.frame`.

Before running the script change the directory to where you have stored the 'GDP_World_Bank.csv'.
```{r}
# dir.string <- "C:/Users/SP4/Documents/Exploring and Analyzing Data/W203 Async/W203 Week 5/Lab 1"

dir.string <- "C:/Users/jhabib/documents/r/W203/W203 Week 5/Lab 1"
setwd(dir.string)

gdp.worldbank <- read.csv("GDP_World_Bank.csv", header = TRUE)
```


`gdp_growth` is the difference between `gdp2012` and `gdp2011`.
```{r}
gdp.worldbank$gdp_growth <- gdp.worldbank$gdp2012 - gdp.worldbank$gdp2011

# gdp.worldbank$gdp_growth
```

We then calculate the mean of `gdp_growth`; we will ingore `NA` values in the mean calculation.
```{r}
gdp.growth.mean <- mean(gdp.worldbank$gdp_growth, na.rm = TRUE)
gdp.growth.mean

```
The calculation above reveals the `gdp.growth.mean` to be `return(gdp.growth.mean)`.

## 11. Create a histogram of `gdp_growth`
We can create a basic histogram using the base R. We will ignore the `NA` values in `gdp_growth`.
```{r}
gdp.worldbank.subset <- gdp.worldbank[!is.na(gdp.worldbank$gdp_growth),c(1, 3, 4, 5)]

hist(gdp.worldbank.subset$gdp_growth, 
     freq = TRUE, 
     breaks = diff(range(gdp.worldbank.subset$gdp_growth))/1e11, 
     xlab = "gdp_growth", 
     ylab = "Frequency", 
     col = "lightblue")
par(new = TRUE)
abline(v = mean(gdp.worldbank.subset$gdp_growth), col="red")
```

We will use the ggplot2 library to create a nicer histogram than the base R package. This time we will plot the density on y-axis.
```{r}
require(ggplot2)

gdp.growth.hist <- ggplot(gdp.worldbank.subset, aes(gdp_growth)) + 
  geom_histogram(na.rm = TRUE, 
                 bins = diff(range(gdp.worldbank.subset$gdp_growth))/1e11, 
                 aes(fill = ..density.., y = ..density..), 
                 col = "blue") + 
  geom_vline(aes(xintercept=mean(gdp.worldbank.subset$gdp_growth), 
                 colour = "red"), 
             size = 1.5) + 
  labs(title="Histogram for GDP Growth") + 
  labs(x="gdp_growth", y="Density")

gdp.growth.hist
```

Let's add a normal curve to our graph to visually see how it fits.
```{r}
gdp.growth.hist + 
  stat_function(fun=dnorm, 
                colour = "red", 
                size = 1, 
                args=list(mean=mean(gdp.worldbank.subset$gdp_growth), 
                          sd=sd(gdp.worldbank.subset$gdp_growth)))
```

This looks like a positively skewed, leptokurtic distribution. Normal distribution does not seem like a good fit. 

Let's create a table with some basic parameters that describe this distribution.
```{r}
gdp.growth.summary.parameter <- c("Mean", "Median", "Minimum", "Maximum", "Standard Deviation", "Skewness", "Kurtosis")
gdp.growth.summary.values <- c(gdp.growth.mean, 
                               median(gdp.worldbank.subset$gdp_growth), 
                               min(gdp.worldbank.subset$gdp_growth), 
                               max(gdp.worldbank.subset$gdp_growth), 
                               sd(gdp.worldbank.subset$gdp_growth), 
                               e1071::skewness(gdp.worldbank.subset$gdp_growth), 
                               e1071::kurtosis(gdp.worldbank.subset$gdp_growth))
gdp.growth.summary <- data.frame(gdp.growth.summary.parameter, gdp.growth.summary.values)
gdp.growth.summary
```


We can use the Jarque-Bera Test from `tseries` to determine goodness of fit.
```{r}
require(tseries)

jarque.bera.test(gdp.worldbank.subset$gdp_growth)

```
It seems that the p-value of 2.2e-16 means that our null-hypothesis of normal distribution is very unlikely.

## 12. Create a `high_growth` boolean variable
We will create a `high_growth` variable that returns 1 if gdp_growth is greater than mean for a country and 0 if not.
```{r}
gdp.worldbank.subset$high_growth <- ifelse(gdp.worldbank.subset$gdp_growth > gdp.growth.mean, 
                                        1, 
                                        0)
high.growth.countries <- sum(gdp.worldbank.subset$high_growth)

low.growth.countries <- length(gdp.worldbank.subset$high_growth) - sum(gdp.worldbank.subset$high_growth)

```
There are 31 countries with high gdp growth by our measure and 142 countries without. These results exclude `NA` values. This makes sense given the distribution of `gdp_growth` because of the positive skewness.

Consider the histogram of `gdp_growth` with resized bins to better show the values on either side of the mean.
```{r}
gdp.growth.hist1 <- ggplot(gdp.worldbank.subset, aes(gdp_growth)) + 
  geom_histogram(na.rm = TRUE, 
                 bins = diff(range(gdp.worldbank.subset$gdp_growth))/1e9, 
                 aes(fill = ..density.., y = ..density..), 
                 col = "blue") + 
  geom_vline(aes(xintercept=mean(gdp.worldbank.subset$gdp_growth), 
                 colour = "red"), 
             size = 0.5) + 
  labs(title="Histogram for GDP Growth") + 
  labs(x="gdp_growth", y="Density")

gdp.growth.hist1
```

# Part 2b. Data Import
I would like to explore whether changes in the Rural Population of a country are correlated to the `gdp_growth` in the same time period. For this purpose I will use 'SP.RUR.TOTL.ZS' indicator from http://data.worldbank.org/indicator/SP.RUR.TOTL.ZS. This indicator shows rural population as a percent of total population by year. We will import data from years 2010, 2011 and 2012 just because our gdp data frame has the same years.

To make things simpler, we will use the `WDI` package that implements an API to obtain data from data.worldbank.org.

```{r}
require(WDI)
rural.pop.worldbank <- WDI(country = "all", indicator = "SP.RUR.TOTL.ZS", 
                           start = 2010, end = 2012, extra = FALSE, cache = NULL)

head(rural.pop.worldbank)
```
We see that `rural.pop.worldbank` is in long format. We will convert it to wide format.
```{r}
require(reshape2)

# We don't need iso2c so let's set it to NULL
rural.pop.worldbank$iso2c <- NULL

# Rename SP.RUR.TOTL.ZS to a more readable rural_pop_percent
names(rural.pop.worldbank)[names(rural.pop.worldbank) == "SP.RUR.TOTL.ZS"] <- "rural_pop_percent"

# Convert the data to wide format
rural.pop.wide <- dcast(melt(rural.pop.worldbank, 
                             id.vars = c("country", "year")), country+variable~variable+year)
```

Let's merge our gdp and rural population data frames together by country. We will use our `gdp.worldbank.subset` data frame because we removed the `NA` values of gdp from it.
```{r}
country.data.merged <- merge(gdp.worldbank.subset, 
                             rural.pop.wide, 
                             by.x = "Country", 
                             by.y = "country", 
                             incomparables = NULL)

head(country.data.merged)
```

Next we'll add a new variable to the merged data frame to get the change in percent population from 2011 to 2012.
```{r}
country.data.merged$rural_pop_growth <- country.data.merged$rural_pop_percent_2012 - 
  country.data.merged$rural_pop_percent_2011
```


Let's create a scatter plot of `rural_pop_growth` on y-axis and `gdp_growth` on x-axis to see if there are any trends.
```{r}
rur.growth.by.gdp.growth <- ggplot(country.data.merged, 
                                   aes(x = gdp_growth, y = rural_pop_growth)) + 
  geom_point(na.rm = TRUE)

rur.growth.by.gdp.growth
```
It's interesting to note that there does not appear to be a correlation between changes in rural populationand changes in gdp from 2011 to 2012.

May be we can read the graph better if x-axis showed `gdp_growth` as a percent of `gdp2011`.
```{r}
rur.growth.by.gdp.growth1 <- ggplot(country.data.merged, 
                                   aes(x = gdp_growth/gdp2011*100, y = rural_pop_growth)) + 
  geom_point(na.rm = TRUE) + 
  geom_vline(xintercept = 0, color = "red") + 
  geom_hline(yintercept = 0, color = "red")

rur.growth.by.gdp.growth1
```
This graph is readable but it still does not show any obvious correlation between changes in rural population and `gdp_growth`. It seems though that most countries saw a decrease in rural population percentage regardless of the positive or negative changes in `gdp_growth` from 2011 to 2012.