---
title: "W203 Week 5 Lab 1"
author: "Mohammad Jawad Habib"
date: "February 7, 2016"
output: pdf_document
---


# Part 2a. Variable Manipulations

## 10. Load Data and calculate gdp_growth

We'll set the working directory to where we saved the "GDP_World_Bank.csv" file and then load it as a `data.frame`.

```{r}
setwd("C:/Users/SP4/Documents/Exploring and Analyzing Data/W203 Async/W203 Week 5/Lab 1")

gdp.worldbank <- read.csv("GDP_World_Bank.csv", header = TRUE)
```


`gdp_growth` is the difference between `gdp2012` and `gdp2011`.
```{r}
gdp.worldbank$gdp_growth <- gdp.worldbank$gdp2012 - gdp.worldbank$gdp2011

# gdp.worldbank$gdp_growth
```

We then calculate the mean of `gdp_growth`; we will ingore `NA` values in the mean calculation.
The calculation below reveals the `gdp.growth.mean` to be 717,237,6796.
```{r}
gdp.growth.mean <- mean(gdp.worldbank$gdp_growth, na.rm = TRUE)
gdp.growth.mean

```

## 11. Create a histogram of `gdp_growth`
We will use the ggplot2 library to create a nicer histogram than the base R package. We will ignore the `NA` values.
```{r}
library(ggplot2)

gdp.worldbank.subset <- gdp.worldbank[!is.na(gdp.worldbank$gdp_growth),c(1, 5)]

gdp.growth.hist <- ggplot(gdp.worldbank.subset, aes(gdp_growth)) + 
  geom_histogram(na.rm = TRUE, 
                 bins = 20, 
                 aes(fill = ..count..), 
                 col = "blue") + 
  labs(title="Histogram for GDP Growth") + 
  labs(x="gdp_growth", y="Count") 

gdp.growth.hist

```
